Extending K-Means Clustering with Ptolemy’s Inequality

### 2024-10-02 content/5_results.tex
proofread first part of results

The developed variant of the k-Means algorithm was tested by implementing the classical k-Means algorithm (Lloyd's algorithm), Elkan's algorithm, and the Ptolemy-based extension in Python. The code is available on GitHub~\cite{Averitchev2024kMeansPtolemy}. To provide a fair comparison, a variable was included to count the number of distance calculations for each algorithm. This metric was chosen because distance calculations account for the majority of computing time in k-Means algorithms, offering a clear measure of efficiency regardless of implementation optimizations.

Additionally, two intermediate variants were implemented to assess the separate effects of the lower and upper bounds: one uses Elkan's lower bound with Ptolemy's upper bound, and the other uses Ptolemy's lower bound with Elkan's upper bound.

\subsection{Datasets}

A variety of datasets were used to compare the performance of the algorithms. The datasets are summarized in Table~\ref{tab:datasets}. 
\begin{table}[htbp]
\centering
\caption{Datasets Used in the Experiments}
\label{tab:datasets}
\begin{tabular}{lll}
\hline
\textbf{Dataset} & \textbf{Cardinality} & \textbf{Dimensionality} \\
\hline
\multicolumn{3}{l}{\textit{Fixed-Structure Datasets}} \\
Iris & 150 & 4 \\
Wine & 178 & 13 \\
\hline
\multicolumn{3}{l}{\textit{Variable-Dimension Datasets}} \\
Gaussian Low Dimension & 10,000 & 10 \\
Gaussian Medium Dimension & 10,000 & 100 \\
Gaussian High Dimension & 10,000 & 300 \\
Random Low Dimension & 5,000 & 10 \\
Random Medium Dimension & 5,000 & 100 \\
Random High Dimension & 5,000 & 300 \\
\hline
\end{tabular}
\end{table}

The fixed-structure datasets include the Iris and Wine datasets \cite{pedregosa2011scikit}, which have fixed dimensions and cardinalities and represent different clustering structures. The variable-dimension datasets include Gaussian and Uniform Random datasets, which offer flexibility in terms of dimensionality and cardinality, making them ideal for testing the algorithms under various conditions.

\subsection{Experiments}

The experiments were designed to evaluate the algorithms based on the number of distance calculations, which dominate the computing time in k-Means algorithms. This metric ensures comparability across different implementations and hardware setups. Each algorithm was executed three times on each dataset, with the parameters $k = 3$, $k = 20$, and $k = 100$.

>>>>

\todo{If we claim to have implemented the classic k-means, we also have to show that in our plots.}
Our variant of the k-Means algorithm is tested by implementing the classical k-Means algorithm (Lloyd's algorithm), Elkan's algorithm, and its novel, Ptolemy-based extension in Python.
The code is available on GitHub~\cite{Averitchev2024kMeansPtolemy}.
To provide a fair comparison of the algorithms,
we measure the number of distance calculations instead of wall-clock time.
This metric was chosen as distance calculations account for the majority of computing time in k-Means algorithms, offering a clear measure of efficiency regardless of implementational optimizations.

Additionally, two intermediate variants were implemented to assess the separate effects of the lower and upper bounds:
\emph{Ptolemy\_upper} one uses Elkan's lower bound with Ptolemy's upper bound, and \emph{Ptolemy\_lower} uses Ptolemy's lower bound with Elkan's upper bound.

\subsection{Datasets}
We employ two groups of datasets to compare the different algorithms, summarized in Table~\ref{tab:datasets}:
The two real-world datasets include the Iris and Wine datasets \cite{pedregosa2011scikit}, which have fixed dimensions and cardinalities and represent different clustering structures.
The two synthetic datasets include Gaussian and uniformly randomly generated points, which are used to test for the effects of changing dataset dimensionality.

\begin{table}[bp]
	\centering
	\caption{Datasets Used in the Experiments}
	\label{tab:datasets}
	\begin{tabular}{llSS}
		\toprule
		\textbf{Group}                                    & \textbf{Dataset}          & \textbf{Cardinality} & \textbf{Dimensionality} \\
		\midrule
		\multirow{2}{*}{\textit{fixed structure}}         & Iris                      & 150                  & 4                       \\
		                                                  & Wine                      & 178                  & 13                      \\
		\cmidrule{2-4}
		\multirow{6}{*}{\textit{variable dimensionality}} & Gaussian Low Dimension    & 10000                & 10                      \\
		                                                  & Gaussian Medium Dimension & 10000                & 100                     \\
		                                                  & Gaussian High Dimension   & 10000                & 300                     \\
		                                                  & Random Low Dimension      & 5000                 & 10                      \\
		                                                  & Random Medium Dimension   & 5000                 & 100                     \\
		                                                  & Random High Dimension     & 5000                 & 300                     \\
		\bottomrule
	\end{tabular}
\end{table}

\subsection{Experiments}
The experiments were designed to evaluate the algorithms based on the number of distance calculations, which dominate the computing time in k-Means algorithms. This metric ensures comparability across different implementations and hardware setups. Each algorithm was executed three times on each dataset, with the parameters $k = 3$, $k = 20$, and $k = 100$.
